\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{setspace}
\usepackage{hyperref}

\setstretch{1.1}

\title{Thesis Summary}
\author{Francesco Giuseppe Minisini}
\date{\today}

\begin{document}

\maketitle

\noindent
This thesis has investigated whether, and to what extent, the framework introduced in \emph{Universal Quantum Control through Deep Reinforcement Learning} by Niu, Boixo, Smelyanskiy and Neven (2019) can be regarded as a physically meaningful and empirically credible strategy for robust analog quantum control on superconducting qubits. 
More specifically, the work was structured around a double objective. The first was to reconstruct, as faithfully as possible, the main physical and methodological ingredients of the original framework, including the multilevel gmon model, the leakage-aware UFO objective, and the continuous-action reinforcement-learning formulation based on TRPO. The second was to assess this framework critically through an independent empirical study, with particular attention to its nominal performance, its robustness under stochastic perturbations, and its runtime behavior across a physically relevant family of target gates.

From the theoretical point of view, the thesis first developed the broader control-theoretic and physical background needed to situate the problem properly. In particular, it emphasized that realistic quantum gate design on near-term superconducting hardware is intrinsically a multi-objective problem, in which gate fidelity, runtime, leakage suppression, and robustness cannot be treated independently. This perspective motivated the discussion of weakly anharmonic multilevel qubits, computational and leakage subspaces, the speed--leakage trade-off, and the general role of model-based optimal control and reinforcement learning in pulse-level quantum control. On this basis, the thesis then reconstructed the paper-specific framework in a more explicit and organized form, clarifying the role of the effective gmon Hamiltonian, the time-dependent Schrieffer--Wolff treatment of leakage, the structure of the UFO cost function, and the logic by which the control problem is cast as a sequential decision process.

From the implementation point of view, the thesis provided a coherent simulation pipeline in which the original framework could be studied quantitatively. The controlled dynamics were simulated in a truncated multilevel Hilbert space, the target gates were chosen from the family \(N(\alpha,\alpha,\gamma)\), and the control objective was evaluated through the same general balance of fidelity, leakage, boundary regularity, and runtime that underlies the original UFO proposal. This made it possible to compare different optimization strategies under matched physical assumptions and to interpret the results within a common methodological setting, rather than as isolated numerical exercises.

The empirical part of the thesis then led to a differentiated but overall coherent picture. On the representative single-target benchmark, the direct Adam baseline proved stronger than the nominal TRPO controller in the deterministic regime, showing that reinforcement learning does not automatically dominate a strong direct pulse optimizer on a fixed nominal problem. However, once the analysis was extended to stochastic control perturbations, the noise-optimized TRPO controller emerged as the strongest overall method, achieving both the highest average fidelity and the lowest fidelity variance under noisy evaluation. This result supports one of the main scientific messages of the original paper: robustness against stochastic control errors can be learned more effectively when the relevant perturbation model is already embedded in the training environment. In parallel, the runtime sweep over the family \(N(\alpha,\alpha,\pi/2)\) showed that analog control can remain substantially faster than standard gate synthesis over a broad region of target space, with the strongest gains appearing where the target gate is most naturally aligned with the native interaction structure of the hardware.

The parameter studies also contributed to the final interpretation of the framework. The architecture sweep showed that the performance of the RL controller depends to a non-negligible extent on the actor--critic representation, but also that this dependence is not monotonic: increasing width or depth does not automatically improve the learned control policy. In this respect, the paper architecture \(64\text{-}32\text{-}32\) remained highly competitive and was retained for the main experiments as the most appropriate paper-faithful choice. Likewise, the sensitivity study on the UFO coefficients showed that the original scalarization \((\chi,\beta,\mu,\kappa)=(10,10,0.2,0.1)\) remains close to the best trade-off region in the present implementation, while also confirming that the balance between fidelity, leakage, boundary regularity, and runtime has a physically interpretable influence on the final controller. These studies reinforced the methodological credibility of the empirical chapter by showing that the main results do not rest on arbitrary hidden parameter choices.

Taken together, these results suggest that the main significance of the validated framework lies not in offering a universally dominant nominal optimizer, but in providing a unified and physically well-motivated strategy for robust analog quantum control. The present thesis therefore supports the view that the real strength of the UFO framework is its ability to integrate, within a single control architecture, several competing requirements that are central to realistic superconducting-qubit operation. In this sense, the work contributes both a structured reconstruction of the original framework and a more critical empirical assessment of its range of validity, its practical advantages, and its limitations.

\end{document}
