\chapter{Conclusion}
\section{Summary of the thesis}
\label{sec:summary_thesis}

This thesis has investigated whether, and to what extent, the framework introduced in \emph{Universal Quantum Control through Deep Reinforcement Learning} can be regarded as a physically meaningful and empirically credible strategy for robust analog quantum control on superconducting qubits \cite{Niu2019}. More specifically, the work was structured around a double objective. The first was to reconstruct, as faithfully as possible, the main physical and methodological ingredients of the original framework, including the multilevel gmon model, the leakage-aware UFO objective, and the continuous-action reinforcement-learning formulation based on TRPO. The second was to assess this framework critically through an independent empirical study, with particular attention to its nominal performance, its robustness under stochastic perturbations, and its runtime behavior across a physically relevant family of target gates.

From the theoretical point of view, the thesis first developed the broader control-theoretic and physical background needed to situate the problem properly. In particular, it emphasized that realistic quantum gate design on near-term superconducting hardware is intrinsically a multi-objective problem, in which gate fidelity, runtime, leakage suppression, and robustness cannot be treated independently. This perspective motivated the discussion of weakly anharmonic multilevel qubits, computational and leakage subspaces, the speed--leakage trade-off, and the general role of model-based optimal control and reinforcement learning in pulse-level quantum control. On this basis, the thesis then reconstructed the paper-specific framework in a more explicit and organized form, clarifying the role of the effective gmon Hamiltonian, the time-dependent Schrieffer--Wolff treatment of leakage, the structure of the UFO cost function, and the logic by which the control problem is cast as a sequential decision process.

From the implementation point of view, the thesis provided a coherent simulation pipeline in which the original framework could be studied quantitatively. The controlled dynamics were simulated in a truncated multilevel Hilbert space, the target gates were chosen from the family \(N(\alpha,\alpha,\gamma)\), and the control objective was evaluated through the same general balance of fidelity, leakage, boundary regularity, and runtime that underlies the original UFO proposal. This made it possible to compare different optimization strategies under matched physical assumptions and to interpret the results within a common methodological setting, rather than as isolated numerical exercises.

The empirical part of the thesis then led to a differentiated but overall coherent picture. On the representative single-target benchmark, the direct Adam baseline proved stronger than the nominal TRPO controller in the deterministic regime, showing that reinforcement learning does not automatically dominate a strong direct pulse optimizer on a fixed nominal problem. However, once the analysis was extended to stochastic control perturbations, the noise-optimized TRPO controller emerged as the strongest overall method, achieving both the highest average fidelity and the lowest fidelity variance under noisy evaluation. This result supports one of the main scientific messages of the original paper: robustness against stochastic control errors can be learned more effectively when the relevant perturbation model is already embedded in the training environment. In parallel, the runtime sweep over the family \(N(\alpha,\alpha,\pi/2)\) showed that analog control can remain substantially faster than standard gate synthesis over a broad region of target space, with the strongest gains appearing where the target gate is most naturally aligned with the native interaction structure of the hardware.

The parameter studies also contributed to the final interpretation of the framework. The architecture sweep showed that the performance of the RL controller depends to a non-negligible extent on the actor--critic representation, but also that this dependence is not monotonic: increasing width or depth does not automatically improve the learned control policy. In this respect, the paper architecture \(64\text{-}32\text{-}32\) remained highly competitive and was retained for the main experiments as the most appropriate paper-faithful choice. Likewise, the sensitivity study on the UFO coefficients showed that the original scalarization \((\chi,\beta,\mu,\kappa)=(10,10,0.2,0.1)\) remains close to the best trade-off region in the present implementation, while also confirming that the balance between fidelity, leakage, boundary regularity, and runtime has a physically interpretable influence on the final controller. These studies reinforced the methodological credibility of the empirical chapter by showing that the main results do not rest on arbitrary hidden parameter choices.

Taken together, these results suggest that the main significance of the validated framework lies not in offering a universally dominant nominal optimizer, but in providing a unified and physically well-motivated strategy for robust analog quantum control. The present thesis therefore supports the view that the real strength of the UFO framework is its ability to integrate, within a single control architecture, several competing requirements that are central to realistic superconducting-qubit operation. In this sense, the work contributes both a structured reconstruction of the original framework and a more critical empirical assessment of its range of validity, its practical advantages, and its limitations.

% %\textbf{TODO:} Inserire qui, se opportuno, una frase finale che richiami in modo sintetico i risultati definitivi dello studio sulle architetture delle policy/value network.

% %\textbf{TODO:} Valutare se aggiungere un breve richiamo conclusivo ai risultati definitivi della sensitivity analysis dei pesi della UFO cost function, solo se tali risultati verranno ritenuti rilevanti per il messaggio complessivo della tesi.


\section{Limitations and future work}
\label{sec:limitations_future_work}

Despite the encouraging results obtained in this thesis, the present study also has a number of clear limitations that should be stated explicitly. First, the entire analysis remains simulation-based. Although the physical model is designed to stay close to a realistic superconducting-qubit setting, the framework has not been validated directly on hardware. As a consequence, the results should be interpreted as evidence about the behavior of the method within a physically motivated numerical environment, rather than as a direct demonstration of experimental performance. In particular, real devices may introduce additional distortions, calibration drifts, transfer-function nonlinearities, spurious modes, and measurement constraints that are only partially captured by the present simulation pipeline.

A second limitation concerns the modeling assumptions adopted throughout the work. The dynamics are simulated in a truncated multilevel Hilbert space with three local levels per mode, under the effective rotating-wave description used in the original framework. While this is sufficient to capture the dominant leakage channels and to preserve the main physical structure of the problem, it remains an approximation. Higher excited states, more detailed control-transfer effects, and more complete open-system dynamics may all become relevant in experimentally realistic regimes. Similarly, the stochastic robustness study has been carried out primarily under a Gaussian control-noise model, with fluctuations injected into the control amplitudes and into the anharmonicity-related parameter. This is physically meaningful and consistent with the original paper, but it does not exhaust the broader space of imperfections affecting superconducting hardware.

A third limitation concerns the optimization comparison itself. Although the thesis includes a meaningful non-RL baseline through direct Adam-based pulse optimization, and therefore does not compare the RL controller against a weak reference, the space of alternative control methods is much broader. In particular, the present study does not provide a fully matched comparison against more refined families of quantum optimal-control approaches, such as GRAPE/Krotov variants with automatic differentiation, differentiable-programming-based controllers, or more recent hybrid model-based / learning-based strategies. For this reason, the conclusions of the thesis should be understood as a comparison between a specific RL-based framework and a specific strong direct-optimization baseline, rather than as a definitive ranking of all available control methodologies.

A further limitation is that some parts of the empirical assessment remain sensitive to implementation-level choices. As discussed in the parameter-study section, quantities such as the training horizon, stopping criterion, curriculum schedule, policy/value architecture, and the scalarization induced by the UFO weights can affect the quantitative behavior of the learned controller. The architecture sweep and the UFO-weight sensitivity analysis suggest that the main qualitative conclusions are stable, but they also make clear that some numerical details should be interpreted with caution. In particular, the strongest conclusions of the work concern the qualitative role of stochastic training and the existence of a runtime advantage in hardware-aligned regions of target space, rather than the exact numerical value of every observed improvement.

These limitations are also useful for positioning the present thesis within the broader evolution of the field after the publication of Niu \emph{et al.} \cite{Niu2019}. In retrospect, one of the main impacts of that paper was not simply to propose another optimizer, but to help establish reinforcement learning as a credible language for formulating robust quantum-control problems with sequential structure, leakage-aware costs, and stochastic perturbations. Subsequent work has broadly confirmed this intuition, but has also refined it in important ways. On the one hand, later research has moved toward genuinely \emph{model-free} or hardware-interacting RL schemes, precisely to reduce model bias and adapt the control policy directly to the device on which it is deployed \cite{Sivak2022}. On the other hand, several works have shown that pure model-free RL is often too sample-inefficient for realistic deployment, and have therefore developed more sample-efficient \emph{model-based} or hybrid strategies that learn or refine an internal dynamical model while preserving part of the flexibility of reinforcement learning \cite{Khalid2023}. In parallel, newer approaches have incorporated stronger physical priors, explicit signal constraints, or demonstration-based warm starts in order to stabilize training and improve scalability. In this sense, the trajectory of the literature since 2019 suggests that the original UFO framework was influential less as a final algorithmic solution than as an early proof that robust analog quantum control could profitably be cast as a learning problem.

This perspective also helps clarify where the frontier currently lies. The literature no longer treats pulse optimization, calibration, compilation, and benchmarking as entirely separate layers. Instead, recent work increasingly points toward integrated workflows in which data-driven characterization, pulse optimization, and hardware validation are tightly coupled. Examples include machine-learning-assisted characterization followed by optimal control on superconducting qubits, reinforcement-learning-based quantum compilation directly on superconducting hardware \cite{Wang2024}, and closed-loop protocols in which the quality of the optimized control is evaluated through gate-set-aware metrics rather than through idealized simulation-only fidelities \cite{Vetter2024,Glaser2025}. Even more broadly, robust-control research is now extending beyond closed-system unitary optimization toward explicitly open-system and experimentally validated settings. Taken together, these developments suggest that the field is progressively moving away from purely black-box deep RL and toward more structured, physics-aware, and experimentally anchored control pipelines.

Against this background, several natural directions for future work emerge from the present thesis. A first priority would be to extend the robustness study beyond the Gaussian perturbation model used here. In particular, it would be scientifically valuable to test whether the robustness learned under stochastic amplitude noise generalizes to quasi-static drift, temporally correlated noise, parameter mismatch, pulse distortions, and simple open-system decoherence channels. Such an extension would make it possible to distinguish more clearly between robustness that is specific to the training distribution and robustness that reflects a broader structural regularity of the learned control strategy.

A second important direction concerns the comparison with more modern hybrid quantum-control methods. The present results already show that direct gradient-based optimization remains very strong in the nominal single-target setting. A more systematic study could therefore compare RL-based control, GRAPE-like methods with automatic differentiation, differentiable-programming controllers, and recent model-based RL approaches under fully matched assumptions, keeping fixed the Hamiltonian model, truncation level, control bounds, runtime penalty, and robustness metrics. This would help clarify more sharply which part of the observed performance comes from the physical objective itself, which part comes from the optimization strategy, and which part comes from the interaction between the two.

A third natural direction is to move from simulation-only robust learning toward \emph{hardware-adaptive} control. One possible route, strongly suggested by recent literature, would be to use the present framework to generate physically informed initial pulses and then refine them through closed-loop calibration on device. In such a workflow, the RL controller would no longer be the final optimizer in isolation, but the first stage of a hybrid procedure combining model-based structure with experimental feedback. This would be particularly attractive for superconducting platforms, where small unmodeled distortions can substantially affect the final pulse quality and where recent studies indicate that closed-loop optimization can already achieve very strong improvements in two-qubit gate performance \cite{Glaser2025}.

A fourth promising direction concerns \emph{sample efficiency and transfer}. One of the most persistent challenges for RL-based quantum control remains the cost of collecting enough trajectories to learn a useful policy. This thesis already used curriculum-style transfer across nearby targets in the runtime study, and the results suggest that cross-target continuity is indeed exploitable. Future work could therefore formalize this idea more systematically through transfer learning across gate families, pretraining on analytically simpler tasks, or reinforcement learning from demonstrations generated by strong classical optimizers. In this respect, recent demonstration-guided and model-based approaches offer a particularly promising bridge between the flexibility of RL and the efficiency of more traditional control pipelines \cite{Khalid2023}.

A fifth direction concerns the control objective itself. The UFO cost function remains one of the conceptually strongest ingredients of the original framework because it unifies fidelity, leakage, boundary regularity, and runtime in a single physically interpretable scalarization. At the same time, the sensitivity study in this thesis suggests that the quantitative behavior of the controller still depends in a nontrivial way on how these terms are weighted. A natural continuation would therefore be to move from a fixed scalarization toward adaptive or context-dependent cost balancing, for example by making the weights depend on the target gate family, on the noise regime, or on experimentally measured sensitivities. Such a direction would connect the present work to the broader modern trend toward task-aware and gate-set-aware control evaluation, where the performance of a pulse is judged not only in isolation but also in terms of how it behaves inside a larger computational workflow \cite{Vetter2024}.

Finally, a broader long-term direction would be to connect pulse-level robust control more directly to the higher architectural layers of quantum computing. Recent work suggests that learning-based methods are beginning to influence not only pulse optimization, but also compilation, calibration, feedback, and architecture-level co-design \cite{Wang2024}. In that wider picture, the most interesting role of frameworks such as the one studied in this thesis may not be to replace all traditional control methods, but to serve as one module within a larger adaptive stack, where device characterization, pulse shaping, circuit compilation, and hardware benchmarking inform one another continuously.

In summary, the present thesis supports reinforcement-learning-based robust analog quantum control as a physically meaningful and promising strategy, but it also shows that its strengths are most convincing when interpreted with appropriate precision. Its main value lies in the robust, hardware-aware, and multi-objective regime, not in a universal nominal dominance over all alternatives. The most promising future developments therefore seem likely to come not from scaling the original framework unchanged, but from integrating its main insights with the directions that the field has taken since its publication: more physics-aware learning, more sample-efficient hybridization, more hardware-in-the-loop validation, and a broader treatment of robustness beyond idealized closed-system settings.

% %\textbf{TODO:} Inserire qui una breve limitazione finale legata ai risultati definitivi dello studio sulle architetture delle policy/value network, una volta completata la versione conclusiva di quella analisi.

% %\textbf{TODO:} Valutare se aggiungere qui una frase specifica sui possibili sviluppi futuri relativi alla calibrazione o reinterpretazione dei pesi della UFO cost function, dopo il consolidamento definitivo della sensitivity analysis.

%\textbf{TODO:} Da controllare bibliografia etc

\newpage
\section{Acknowledgements}

I would like to express my deepest gratitude to my supervisors \emph{Prof. Morten Hjorth-Jensen} and \emph{Prof. Davide Emilio Galli}, whose guidance and support have extended far beyond the development of this thesis. Their encouragement, along with other professors like \emph{Lidia Dell'Asta} has played an important role not only in this work, but also in shaping my broader academic path and future aspirations.

I am also sincerely grateful to \emph{Luca Miceli}, whose mentorship introduced me more closely to the world of research and gave me the opportunity to live experiences that profoundly strengthened my enthusiasm for this way of thinking and living science. I would also like to thank Teresa Ghirlandi and Marisa Lonocce, as well as the friends and colleagues who supported me throughout my journey.

My most heartfelt thanks go to my family: to my parents and my sister for their constant love and support, and in a special way to my cousins \emph{Edoardo} and \emph{Vittorio Mocchi} and to my uncle \emph{Antonio Mocchi}, whose encouragement, belief in me, and concrete support were especially important in shaping my ambition, my motivation, and my university path. Finally, I wish to thank \emph{Margherita Frova} for her closeness, affection, and constant presence throughout this period.
